---
title: "Scran_method"
output: html_document
date: "2026-08-23"
---

# Appendix Y: scRNA-seq Analytical Pipeline, Normalization, and Spatial Mapping

This section outlines the core analytical pipeline applied to the *Danio rerio* 10 dpf dataset. The process encompasses mathematical filtering, transcriptomic normalization via the `scran` algorithm, variance modeling for the extraction of highly variable genes (HVGs), dimensionality reduction, and targeted spatial visualization of functional gene modules.

## 1. Quality Assessment and Mathematical Filtering

Initial cell viability is visually assessed before applying strict mathematical thresholds to isolate true biological signals from technical artifacts (such as empty droplets, doublets, or apoptotic cells).

```{r quality_filtering, message=FALSE, warning=FALSE}
# Load required analytical packages
library(Seurat)
library(scran)
library(SingleCellExperiment)

# Import the previously initialized Seurat object
seurat_10dpf <- readRDS("seurat_10dpf_backup.rds")

# Generate and export violin plots for baseline quality control visualization
plot_vln <- VlnPlot(seurat_10dpf, 
        features = c("nFeature_RNA", "pct_counts_mt"), 
        group.by = "fish", 
        pt.size = 0.1)

pdf("VlnPlot.RNA_pct_counts.pdf", width = 12, height = 8)
print(plot_vln)
dev.off()

# Apply strict biological thresholds (Cut-offs) to filter out low-quality cells
# Limits are set to exclude empty droplets/debris (< 200), doublets (> 4000), and dying cells (mt > 5%)
seurat_filtre <- subset(seurat_10dpf, 
                        subset = nFeature_RNA > 200 & 
                                 nFeature_RNA < 4000 & 
                                 pct_counts_mt < 5)

# Verify the matrix dimensions to evaluate the impact of the filtering step
print("Matrix dimensions after filtering:")
dim(seurat_filtre)
```

##2. Normalization and Variance Modeling (scran Integration)
To accurately compute biological variance, the data is converted and processed using the scran package. This approach utilizes cellular pre-pooling (deconvolution) to compute robust size factors, followed by a mathematical modeling of the mean-variance relationship to extract Highly Variable Genes (HVGs).

```{r quality_filtering, message=FALSE, warning=FALSE}
# Convert the filtered Seurat object to a SingleCellExperiment (SCE) framework for scran compatibility
sce_10dpf <- as.SingleCellExperiment(seurat_filtre)

# Perform cellular pre-clustering to calculate accurate size factors
set.seed(100)
clusters_q <- quickCluster(sce_10dpf)

# Deconvolution: Compute size factors to correct for sequencing depth biases
sce_10dpf <- computeSumFactors(sce_10dpf, cluster = clusters_q)

# Apply log-normalization to stabilize the variance across the dataset
sce_10dpf <- logNormCounts(sce_10dpf)
summary(sizeFactors(sce_10dpf))

# Model the intracellular variance using biological replicates (fish) as a blocking factor
dec_10dpf <- modelGeneVar(sce_10dpf, block = sce_10dpf$fish)

# Extract trend curve parameters for the first block (replicate 1) for visualization
bloc_1 <- dec_10dpf$per.block[[1]] courbe_poisson_1 <- metadata(bloc_1)$trend

# Generate and export a scatter plot illustrating the technical noise model
pdf("Scran.Technical_Noise.pdf", width = 12, height = 8)
plot(dec_10dpf$mean, dec_10dpf$total, 
     xlab = "Mean expression (log)", 
     ylab = "Total variance", 
     main = "scran Model - Technical Noise",
     pch = 16, cex = 0.5, col = "darkgrey")
curve(courbe_poisson_1(x), col = "blue", add = TRUE, lwd = 2)
dev.off()

# Extract the top 10% of genes exhibiting the highest biological variance (HVGs)
hvg_10dpf <- getTopHVGs(dec_10dpf, prop = 0.1)
print(paste("Number of isolated HVGs:", length(hvg_10dpf)))

# Re-inject the scran-normalized mathematical matrix directly into the Seurat object
matrice_scran <- logcounts(sce_10dpf)
seurat_filtre[["RNA"]]$data <- matrice_scran
```

##3. Dimensionality Reduction (PCA)
With the biologically relevant HVGs identified, the data is scaled and subjected to Principal Component Analysis (PCA) to reduce computational complexity and highlight the primary axes of transcriptomic variation.

```{r quality_filtering, message=FALSE, warning=FALSE}
# Assign the previously extracted HVGs as the exclusive variable features in Seurat
VariableFeatures(seurat_filtre) <- hvg_10dpf

# Center and scale the data (applied solely to HVGs to optimize computational efficiency)
seurat_filtre <- ScaleData(seurat_filtre, features = VariableFeatures(seurat_filtre))

# Execute Principal Component Analysis (PCA)
seurat_filtre <- RunPCA(seurat_filtre, features = VariableFeatures(seurat_filtre))

# Visualize and export the Elbow Plot to determine the optimal number of principal components
elbow  <- ElbowPlot(seurat_filtre, ndims = 40)
pdf("ElbowPlot.pdf", width = 12, height = 8)
print(elbow)
dev.off()
```

##4. Functional Enrichment Analysis (Gene Ontology)
The isolated biological signal (HVGs) is evaluated against the Danio rerio reference genome to identify significantly over-represented biological processes using Gene Ontology (GO).

```{r quality_filtering, message=FALSE, warning=FALSE}
# Load libraries required for genome annotation and functional enrichment
library(clusterProfiler)
library(org.Dr.eg.db)
library(ggplot2)

# Extract the HVG list
hvg_genes <- VariableFeatures(seurat_filtre)

# Perform Gene Ontology (GO) enrichment analysis targeting Biological Processes (BP)
ego_hvg <- enrichGO(
  gene          = hvg_genes,
  keyType       = "SYMBOL",
  OrgDb         = org.Dr.eg.db,
  ont           = "BP", 
  pAdjustMethod = "BH", 
  universe      = dec_10dpf@rownames,
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05
)

# Visualize and export the 15 most significant functional groups
GO_Plot <- dotplot(ego_hvg, showCategory = 15) + ggtitle("Biological functions for the HVG genes")
pdf("GOPlot_HVG_genes.pdf", width = 12, height = 8)
print(GO_Plot)
dev.off()
```

##5. Multi-Module Analysis and Spatial Mapping (UMAP)

Specific functional modules (e.g., Growth, Sensory Perception, Regeneration) are defined programmatically. Cells are then scored based on the expression of these networks, and the resulting biological signatures are projected onto a 2D spatial approximation (UMAP) to evaluate tissue compartmentalization.

```{r quality_filtering, message=FALSE, warning=FALSE}
# Convert the GO enrichment results to a data frame for programmatic mining
ego_df <- as.data.frame(ego_hvg)

# --- Define and Filter the Growth Module ---
growth_terms <- c("developmental growth", "growth")
raw_growth_genes <- ego_df[ego_df$Description %in% growth_terms, "geneID"]
split_growth_genes <- unique(unlist(strsplit(raw_growth_genes, "/")))
matching_growth_genes <- intersect(split_growth_genes, hvg_genes)

# --- Define and Filter the Sensory Module ---
sensory_terms <- c("sensory organ morphogenesis", "retina development in camera-type eye", 
                   "sensory perception", "visual perception", "sensory perception of light stimulus")
raw_sensory_genes <- ego_df[ego_df$Description %in% sensory_terms, "geneID"]
split_sensory_genes <- unique(unlist(strsplit(raw_sensory_genes, "/")))
matching_sensory_genes <- intersect(split_sensory_genes, hvg_genes)

# --- Compute Targeted Module Scores ---
# Scores are added directly to the Seurat object metadata
seurat_filtre <- AddModuleScore(seurat_filtre, features = list(matching_growth_genes), name = "Score_Growth")
seurat_filtre <- AddModuleScore(seurat_filtre, features = list(matching_sensory_genes), name = "Score_Sensory")

# --- Anatomical Distribution Analysis (Violin Plots) ---
vln_growth <- VlnPlot(seurat_filtre, features = "Score_Growth1", group.by = "zebrafish_anatomy_ontology_class", pt.size = 0) + ggtitle("Anatomical Distribution of Growth Score")
vln_sensory <- VlnPlot(seurat_filtre, features = "Score_Sensory1", group.by = "zebrafish_anatomy_ontology_class", pt.size = 0) + ggtitle("Anatomical Distribution of Sensory Score")

pdf("VlnPlot.Growth.pdf", width = 18, height = 8)
print(vln_growth)
dev.off()

pdf("VlnPlot.Sensory.pdf", width = 18, height = 8)
print(vln_sensory)
dev.off()

# --- Latent Space Spatial Visualization (UMAP) ---
# Compute the UMAP projection based on the first 20 principal components
seurat_filtre <- RunUMAP(seurat_filtre, dims = 1:20)

# Map the Growth score onto the UMAP coordinates (cold blue gradient)
umap_growth <- FeaturePlot(seurat_filtre, features = "Score_Growth1", reduction = "umap", cols = c("lightgrey", "blue")) + ggtitle("Expression of Growth Module on UMAP")

# Map the Sensory score onto the UMAP coordinates (deep purple gradient)
umap_sensory <- FeaturePlot(seurat_filtre, features = "Score_Sensory1", reduction = "umap", cols = c("lightgrey", "purple")) + ggtitle("Expression of Sensory Module on UMAP")

pdf("UMAP.Growth.pdf", width = 18, height = 8)
print(umap_growth)
dev.off()

pdf("UMAP.Sensory.pdf", width = 18, height = 8)
print(umap_sensory)
dev.off()
```




