---
title: "Seurat"
output: html_document
date: "2026-08-21"# Appendix X: scRNA-seq Data Import and Seurat Object Initialization

The following script details the computational pipeline used to import raw single-cell transcriptomic data (scRNA-seq) from the *Danio rerio* 10 dpf model. Since the original dataset is provided in a Python-based AnnData format (`.h5ad`), the `reticulate` package is utilized to bridge the Python and R environments. This ensures a lossless conversion of the sparse matrices and metadata prior to their integration into the Seurat analytical ecosystem.

## 1. Environment Setup and Data Loading

Required libraries are loaded to enable data manipulation, single-cell analysis, and Python interoperability. The raw dataset is then imported without automatic conversion to preserve its complex structure.

```{r setup, message=FALSE, warning=FALSE}
# Load required libraries
library(tidyverse)
library(Seurat)
library(reticulate)

# Import required Python modules using reticulate 
ad <- import("anndata", convert = FALSE)
scipy <- import("scipy.sparse", convert = FALSE)

# Load the raw transcriptomic data stored in AnnData format (.h5ad)
adata <- ad$read_h5ad("zf_atlas_10dpf_v1_release.h5ad")
print(adata)
```

## 2. Metadata Extraction and Matrix Transformation
The metadata containing cellular annotations is extracted. Subsequently, the count matrix is transposed. This structural transposition is mandatory because Python (AnnData) stores matrices as (cells x genes), whereas R (Seurat) strictly requires a (genes x cells) format.

```{r setup, message=FALSE, warning=FALSE}
# Extract metadata (cell annotations) and convert to an R data frame
metadata <- py_to_r(adata$obs)
dim(metadata)        
head(metadata[, 1:4])

# Extract and transpose the expression count matrix
X_transposed <- adata$X$transpose()
counts_matrix <- py_to_r(X_transposed)

# Request Python to transform the indices into pure lists, then convert them to R vectors
gene_names <- py_to_r(adata$var_names$to_list())
cell_names <- py_to_r(adata$obs_names$to_list())

# Assign the extracted gene and cell names to the R count matrix
rownames(counts_matrix) <- gene_names
colnames(counts_matrix) <- cell_names
```

## 3. Seurat Object Creation and Verification
With the correctly formatted matrix and metadata, the Seurat object is initialized. A series of exploratory commands are then executed to verify the integrity of the data structure, the distribution of biological replicates, and the sparsity of the transcriptomic signal.

```{r setup, message=FALSE, warning=FALSE}
# Construct the Seurat object 
seurat_10dpf <- CreateSeuratObject(counts = counts_matrix, meta.data = metadata)

# Verify the Seurat object dimensions and size
print(seurat_10dpf)

# Inspect the first few rows and columns of the matrix and metadata for quality control
counts_matrix[1:10, 1:5]
colnames(metadata)
head(rownames(metadata), 10)

# Explore the available metadata categories provided by the dataset authors
colnames(seurat_10dpf@meta.data)

# Quantify the number of cells originating from each individual biological replicate (fish)
table(seurat_10dpf$fish)

# List all annotated tissues present in the dataset and quantify the cellular population for each
table(seurat_10dpf$zebrafish_anatomy_ontology_class)

# Generate a cross-tabulation to evaluate the distribution of cell types across biological replicates
table(seurat_10dpf$fish, seurat_10dpf$zebrafish_anatomy_ontology_class)

# Verify the computational format of the count matrix (ensuring it is a sparse matrix structure)
class(seurat_10dpf[["RNA"]]$counts)

# Inspect raw UMI counts and potential technical noise for a broadly expressed gene (actb1)
seurat_10dpf[["RNA"]]$counts["actb1", 1:5]

# Inspect counts for a highly specific gene (rho) to evaluate biological specificity and sparsity
seurat_10dpf[["RNA"]]$counts["rho", 1:5]```

